---
title: "hiermod"
output: html_document
date: "2026-07-07"
---

# Setup 
with rarefied diversity 
```{r}
pacman::p_load(tidyverse, rethinking)
analysis_shared_path <- '~/USherbrooke/Isabelle Laforest-Lapointe - MICROBIAL_ECOLOGY/INTERANNUAL_22_23_24/ANALYSIS'
```


```{r, eval = FALSE, include=FALSE}
# Temporary
ps_ls <- read_rds(file.path(analysis_shared_path, 'R_data/ps_list_ITS.rds'))
div.ITS <- mgx.tools::rarefy_diversity(ps_ls$raw_its,depth = 4500)
write_rds(div.ITS, file.path(analysis_shared_path, 'R_data/diversity/ITS_div.rds'))
```

```{r}
div.ITS <- read_rds(file.path(analysis_shared_path, 'R_data/diversity/ITS_div.rds'))
```

## Simple model: effect of Management on diversity

DAG_1 : Management ---> Diversity

### Simulate diversity from management type 

We create a very simple generative model where the mean diversity is determined by management type (conventional or organic)

```{r}
sim_div_man <- function(man, a){
  N <- length(man)
  D <- a[man] + rnorm(N, 0, 5)
  data.frame(man, D)
}

# By design for two categories so we can use rbern()
sim_div_man(man = rbern(100)+1, a = c(5,10)) %>% head()

```


$D_i \sim Normal(\mu_i, \sigma)$  
$\mu_i = \alpha_{M[i]}$  

Priors:  
$\alpha_j\sim Normal(20,15)$  
$\sigma \sim Exponential(5)$

### Testing
Total causal effect of management on diversity?
Using 100 samples in each type, compute the mean difference in

```{r}

sim_con <- sim_div_man(man = rep(1,100), a = c(5,10))
sim_org <- sim_div_man(man = rep(2,100), a = c(5,10))

mean(sim_org$D - sim_con$D)

```

We are god, we measured the effect of management on diversity!

### Estimator

Simulate a sample.

```{r}
# OBSERVE A SAMPLE :
man <- rbern(100)+1
dat_sim <- sim_div_man(man, a = c(5,10))

# ESTIMATE THE POSTERIOR:
model_MD <- quap(
  alist(
    D ~ dnorm(mu, sigma),
    mu <- a[man],
    a[man] ~ dnorm(20,15),
    sigma ~ dexp(5)
  ), data = dat_sim)

precis(model_MD, depth = 2)

```
difference is about 5, which makes sense because that's what we made it to be.


### Analyze the real sample
If the assumptions about the DAG and the data generation process, 


```{r}

dat_real <- list(
  D = div.ITS$alpha$Hill_1,
  man = ifelse( div.ITS$alpha$Management == 'CONV', 1, 2))

model_MD_real <- quap(
  alist(
    D ~ dnorm(mu, sigma),
    mu <- a[man],
    a[man] ~ dnorm(20, 15),
    sigma ~ dexp(5)
  ), data = dat_real)

precis(model_MD_real, depth = 2)

```


### Compare posterior distributions

The difference in mean diversity is not *in* the posterior; it needs to be computed by contrasting draws from the posterior. 

```{r}
post <- extract.samples(model_MD_real)

# POSTERIOR MEAN DIVERSITY
dens (post$a[,1], lwd = 3, col = 2, 
      xlim = c(floor(min(post$a)),ceiling(max(post$a))),
      xlab = "Posterior mean diversity (species)")
dens (post$a[,2], lwd=3, col = 4, add = TRUE)

# POSTERIOR DIVERSITY DISTRIBUTIONS
D1 <- rnorm(1000, post$a[,1], post$sigma)
D2 <- rnorm(1000, post$a[,2], post$sigma)

dens( D1, xlim = c(floor(min(c(D1,D2))), ceiling(max(c(D1,D2)))),
      lwd=3, col=2) # red = conventional
dens( D2, lwd=3, col = 4, add = TRUE)
```


We still have to compute the *contrasts*, because the overlap is not indicative of the difference. We are looking for the posterior distribution of the DIFFERENCES between our categories.

* It is never legitimate to compare overlap in parameter distributions! *

### Compute contrasts 

_Differences in posterior distributions._ 

```{r}

str(post)
```

Compute the contrast using the difference in the two variables in the posterior distributions.
```{r}
# Contrast in MEANS
mu_contrast <- post$a[,2] - post$a[,1]

dens( mu_contrast, lwd = 3, col = 1, 
      xlab = "Posterior mean diversity contrast (effective number of species)")

```

What about the contrast in samples? What is the contrast between those distributions

```{r}
# Contrast in SAMPLE DIVERSITY
D_contrast <- D2 - D1
dens(D_contrast, lwd = 3, col = 1, 
     xlab = 'Posterior diversity contrast')
abline(v=0, col=2, lwd = 3)

# Proportion above/below zero:
 # proportion of organic samples expected to be more diverse than conventional?
sum (D_contrast>0)/1000

# or opposite:
sum (D_contrast<0)/1000

```

Interpretation: if you randomly select an organic and a conventional sample from the population, 62% of the time organic will have higher diversity (according to this flawed model).


