---
title: "hiermod"
output: html_document
date: "2026-07-07"
---

# Setup 
with rarefied diversity 
```{r}
pacman::p_load(tidyverse, rethinking)
analysis_shared_path <- '~/USherbrooke/Isabelle Laforest-Lapointe - MICROBIAL_ECOLOGY/INTERANNUAL_22_23_24/ANALYSIS'
```

```{r}
div.ITS <- read_rds(file.path(analysis_shared_path, 'R_data/diversity/ITS_div.rds'))
```

DAG_1 : Management ---> Diversity

## Which distribution would be appropriate?

```{r}
div.ITS$alpha %>% 
  pull(Hill_1) %>% sort() %>% 
  hist(breaks = 30)

```
Simulate a log-normal for 240 samples, find parameters that fit the scale
```{r}
rlnorm(240, meanlog = log(10), sdlog = 0.5) %>% 
  hist(breaks = 30) 
```


### Simulate diversity from management type 

We create a less simple generative model where the mean diversity is determined by management type (conventional or organic). 

For this, we use a shifted lognormal to lower-bound values at 1, because Hill numbers can't go below this. 

**Warning!** Shifting will require transforming the outcome log(D - 1) as the modeled variable, then back-transforming) or writing a custom likelihood, when fitting real D values.

```{r}
# ---- Lognormal, shifted to support >= 1, per-category mean + CV ----
sim_div_man_lnorm <- function(management, mean_, cv_){
  N      <- length(management)
  m      <- mean_[management] - 1  # wew want E[D] = mean_ and D = 1 + ...
  cv     <- cv_[management]
  sigma2 <- log(1 + cv^2)          # variance on log-scale; cv is scale-free
  mu     <- log(m) - sigma2/2
  D      <- 1 + rlnorm(
    N, meanlog = mu, 
    sdlog = sqrt(sigma2)) # because rnorm needs a sd
  data.frame(management, D)
}

# Test it 
sim_div_man_lnorm(management = rbern(100)+1, mean_ = c(5,10), cv_ = c(0.5, 0.3))

```

But the mean and cv values are arbitrary; we want to see what happens over a large variation set for those two parameters, so we do SBC.

### Simulated parameter recovery check

```{r}
management <- rbern(100) + 1
sim <- sim_div_man_lnorm(management, mean_ = c(5, 10), cv_ = c(0.3, 0.3))
sim$Dshift <- sim$D - 1   # shift back off, since quap needs the lognormal's own support

model_MD_lnorm <- quap(
  alist(
    Dshift ~ dlnorm(mu, sigma),
    mu <- loga[management],
    loga[management] ~ dnorm(1.5, 1),   # prior on LOG scale now, not raw scale
    sigma ~ dexp(1)
  ),
  data = sim
)

precis(model_MD_lnorm, depth = 2)
```

#### Rescale the contrasts
Since loga is on the log scale, a[1] and a[2] from precis() aren't on D's scale. Back-transform through the lognormal mean formula (E[X] = exp(mu + sigma^2/2)), then add the shift back:

```{r}
post <- extract.samples(model_MD_lnorm)
mean1 <- 1 + exp(post$loga[,1] + post$sigma^2 / 2)
mean2 <- 1 + exp(post$loga[,2] + post$sigma^2 / 2)
contrast <- mean2 - mean1
mean(contrast); PI(contrast); hist(contrast)
```



### Simulation-based calibration (SBC)

```{r}
n_sbc <- 500
N_per_group <- 120  

results <- vector("list", n_sbc)

for (i in seq_len(n_sbc)) {
  
  # 1. Our priors:
  mean1 <- rlnorm(1, meanlog = log(5), sdlog = 0.4)   # tune to your domain
  mean2 <- rlnorm(1, meanlog = log(5), sdlog = 0.4)
  cv1   <- runif(1, 0.1, 0.5)
  cv2   <- runif(1, 0.1, 0.5)
  
  # 2. simulate data under those true params
  man <- c(rep(1, N_per_group), rep(2, N_per_group))
  sim <- sim_div_man_lnorm(man, mean_ = c(mean1, mean2), cv_ = c(cv1, cv2))
  
  # 3. fit your Bayesian model to sim (placeholder)
  # fit <- brm(D ~ 0 + factor(man), family = ..., data = sim)
  
  # 4. store true contrast + posterior draws of contrast for later rank checks
  true_contrast <- mean2 - mean1
  post_contrast <- as_draws_df(fit)$b_man2 - as_draws_df(fit)$b_man1
  results[[i]] <- list(
    true = true_contrast, rank = sum(post_contrast < true_contrast))
}

results
```


### Estimator

Simulate a sample.

```{r}
# OBSERVE A SAMPLE :
management <- rbern(100)+1
dat_sim <- sim_div_man(management, a = c(5,10))

# ESTIMATE THE POSTERIOR:
model_MD <- quap(
  alist(
    D ~ dnorm(mu, sigma),
    mu <- a[management],
    a[management] ~ dnorm(20,15),
    sigma ~ dexp(5)
  ), data = dat_sim)

precis(model_MD, depth = 2)

```
difference is about 5, which makes sense because that's what we made it to be.


### Analyze the real sample
If the assumptions about the DAG and the data generation process, 


```{r}

dat_real <- list(
  D = div.ITS$alpha$Hill_1,
  management = ifelse( div.ITS$alpha$Management == 'CONV', 1, 2))

model_MD_real <- quap(
  alist(
    D ~ dnorm(mu, sigma),
    mu <- a[management],
    a[management] ~ dnorm(20, 15),
    sigma ~ dexp(5)
  ), data = dat_real)

precis(model_MD_real, depth = 2)

```


### Compare posterior distributions

The difference in mean diversity is not *in* the posterior; it needs to be computed by contrasting draws from the posterior. 

```{r}
post <- extract.samples(model_MD_real)

# POSTERIOR MEAN DIVERSITY
dens (post$a[,1], lwd = 3, col = 2, 
      xlim = c(floor(min(post$a)),ceiling(max(post$a))),
      xlab = "Posterior mean diversity (species)")
dens (post$a[,2], lwd=3, col = 4, add = TRUE)

# POSTERIOR DIVERSITY DISTRIBUTIONS
D1 <- rnorm(1000, post$a[,1], post$sigma)
D2 <- rnorm(1000, post$a[,2], post$sigma)

dens( D1, xlim = c(floor(min(c(D1,D2))), ceiling(max(c(D1,D2)))),
      lwd=3, col=2) # red = conventional
dens( D2, lwd=3, col = 4, add = TRUE)
```


We still have to compute the *contrasts*, because the overlap is not indicative of the difference. We are looking for the posterior distribution of the DIFFERENCES between our categories.

* It is never legitimate to compare overlap in parameter distributions! *

### Compute contrasts 

_Differences in posterior distributions._ 

```{r}

str(post)
```

Compute the contrast using the difference in the two variables in the posterior distributions.
```{r}
# Contrast in MEANS
mu_contrast <- post$a[,2] - post$a[,1]

dens( mu_contrast, lwd = 3, col = 1, 
      xlab = "Posterior mean diversity contrast (effective number of species)")

```

What about the contrast in samples? What is the contrast between those distributions

```{r}
# Contrast in SAMPLE DIVERSITY
D_contrast <- D2 - D1
dens(D_contrast, lwd = 3, col = 1, 
     xlab = 'Posterior diversity contrast')

# Proportion above/below zero:
# proportion of organic samples expected to be more diverse than conventional?
sum (D_contrast>0)/1000

# or opposite:
sum (D_contrast<0)/1000

```

Interpretation: if you randomly select an organic and a conventional sample from the population, 62% of the time organic will have higher diversity (according to this flawed model).


